Quantifying Infants' Everyday Restrained Experiences in the Home Using Wearable Inertial Sensors

### 2025-11-21 manuscript_restrained_classification validation.qmd
edited declarations
fixed typo

The past decade has witnessed an uptick of using wearable sensors to record and classify movement categories to enjoy these advantages. The application of wearable sensors progressed from adults [@Arif2015u; @Preece2009n], to children [@Nam2013y; @Ren2016-px; @Stewart2018h], and then to infants [@Airaksinen2020j; @Airaksinen2024z; @Franchak2021k; @Franchak2024e; @Yao2019q]. For infants, there have been established models that classify infants’ body postures in general—including holding and carrying [@Airaksinen2020j; @Franchak2021k; @Franchak2024e], and models that specifically detect holding and carrying periods [@Airaksinen2024z; @Yao2019q]. Insights from prior studies suggest effective practices, such as using multiple sensors and using sensors that record both gyroscope and accelerator data [@Airaksinen2025l; @Franchak2024e].

@@

In the data set, a special infant garment was customized to secure four lightweight wearable sensors (MC10 Biostamp) at four specific locations (i.e., right thigh, left thigh, right ankle, and left ankle). Each sensor collects raw data as accelerometer signals (in x, y, and z orientations) and gyroscope signals (in roll, pitch, and yaw orientations). The sampling rate of 62.5 Hz allows for high resolution motion recording, and the sufficient battery life and built-in storage allow for daylong recording. An action camera fixed on a mini tripod (GoPro HERO9 Black) was used to record synchronized third-person video of the first 1.5 hours into the session as ground truth.

@@

Data collection took place in the family’s home and began with synchronizing the video camera and wearable sensors. The caregiver was instructed to set up the camera to capture the main room and to move it as needed to ensure the infant remained in view. The caregiver put the garment on the infant and positioned the infant in various postures (including holding and sitting in a restraint device) for at least one minute each for model training purpose. Afterwards, the infant continued wearing the garment during their daily routines. The caregiver logged the times when the infant was napping or when the garment was removed (e.g., for diaper changes, baths, or errands).

@@

The third step was to train and validate supervised machine learning models that classify the motion features into restrained status. We applied the random forest algorithm (Breiman, 2001) in modeling using the DecisionTree package (Sadeghi et al., 2022) in Julia. We used a leave-one-out cross-validation approach, where we held out one session at each iteration as a testing set and trained the model using the other sessions, to evaluate how well the model generalized to unseen data. After validation, an optimal window length was picked, and a final model was trained using all the human annotations to process whole-day sensor data of all sessions.

@@

Machine-learning process was conducted in Julia version 1.11.1 using the DecisionTree package (Sadeghi et al., 2022). This reproducible manuscript, including the model evaluation, figures, and statistical analyses, was made with Quarto version 1.5.57 (Posit, PBC, 2024) using R version 4.5.0 [@R-base] with the R packages tidyverse [@R-tidyverse], flextable [@R-flextable], and papaja (Aust & Barth, 2024), as well as the apa quarto template (Schneider, n.d.). Human annotation results, which were used for model training and validation, together with processing scripts are publicly shared on OSF (https://osf.io/a698g/overview?view_only=221d1f2bc8d34b5d9838354224d6da8f).

@@

We first compared model performance to the human annotations of the 1.5-hour video recorded period across all 142 sessions using leave-one-out cross-validation approach. We calculated performance metrics including accuracy, kappa, sensitivity, positive predictive value, F1, and the correlation of overall estimate time of restraint between the model and human annotation. We also examined prediction biases model and how they vary across window lengths. Second, we applied the model with the optimal window length to process infants’ full-day sensor data and examined the model’s convergent validity by replicating a previously-reported effect: that restraint time decreases with age [@Carson2022n].

@@

This work was supported by National Science Foundation Grant BCS #1941449.

>>>>

The past decade has witnessed an uptick of using wearable sensors to record and classify movement categories to enjoy these advantages. The application of wearable sensors progressed from adults [@Arif2015u; @Preece2009n], to children [@Nam2013y; @Ren2016-px; @Stewart2018h], and then to infants [@Airaksinen2020j; @Airaksinen2024z; @Franchak2021k; @Franchak2024e; @Yao2019q]. For infants, there have been established models that classify infants’ body postures in general—including holding and carrying [@Airaksinen2020j; @Franchak2021k; @Franchak2024e], and models that specifically detect holding and carrying periods [@Airaksinen2024z; @Yao2019q]. Insights from prior studies suggest effective practices, such as using multiple sensors and using sensors that record both gyroscope and accelerometer data [@Airaksinen2025l; @Franchak2024e].

@@

In the data set, a special infant garment was customized to secure four lightweight wearable sensors (MC10 Biostamp) at four specific locations (i.e., right thigh, left thigh, right ankle, and left ankle). Each sensor collects raw data as accelerometer signals (in x, y, and z orientations) and gyroscope signals (in roll, pitch, and yaw orientations). The sampling rate of 62.5 Hz allows for high-resolution motion recording, and the sufficient battery life and built-in storage allow for daylong recording. An action camera fixed on a mini tripod (GoPro HERO9 Black) was used to record synchronized third-person video of the first 1.5 hours into the session as ground truth.

@@

Data collection took place in the family’s home and began with synchronizing the video camera and wearable sensors. The caregiver was instructed to set up the camera to capture the main room and to move it as needed to ensure the infant remained in view. The caregiver put the garment on the infant and positioned the infant in various postures (including holding and sitting in a restraint device) for at least one minute each for model training purposes. Afterwards, the infant continued wearing the garment during their daily routines. The caregiver logged the times when the infant was napping or when the garment was removed (e.g., for diaper changes, baths, or errands).

@@

The third step was to train and validate supervised machine learning models that classify the motion features into restrained statuses. We applied the random forest algorithm (Breiman, 2001) in modeling using the DecisionTree package (Sadeghi et al., 2022) in Julia. We used a leave-one-out cross-validation approach, where we held out one session at each iteration as a testing set and trained the model using the other sessions, to evaluate how well the model generalized to unseen data. After validation, an optimal window length was picked, and a final model was trained using all the human annotations to process whole-day sensor data of all sessions.

@@

Machine-learning process was conducted in Julia version 1.11.1 using the DecisionTree package (Sadeghi et al., 2022). This reproducible manuscript, including the model evaluation, figures, and statistical analyses, was made with Quarto version 1.5.57 (Posit, PBC, 2024) using R version 4.5.0 [@R-base] with the R packages tidyverse [@R-tidyverse], flextable [@R-flextable], and papaja (Aust & Barth, 2024), as well as the APA Quarto template (Schneider, n.d.).Human annotation results, processing scripts, and analyses codes are publicly shared on OSF (https://osf.io/a698g/overview?view_only=221d1f2bc8d34b5d9838354224d6da8f).

@@

We first compared model performance to the human annotations of the 1.5-hour video recorded period across all 142 sessions using the leave-one-out cross-validation approach. We calculated performance metrics, including accuracy, kappa, sensitivity, positive predictive value, F1, and the correlation of the overall estimate time of restraint between the model and human annotation. We also examined the prediction biases of the model and how they vary across window lengths. Second, we applied the model with the optimal window length to process infants’ full-day sensor data and examined the model’s convergent validity by replicating a previously-reported effect: that restraint time decreases with age [@Carson2022n].

@@

This work was supported by National Science Foundation Grants BCS #1941449 and BCS #2521429.
